Evolution of the AGN UV luminosity function from redshift 7.5

### 2018-03-12 Paper/paper.tex
Update text

\author[Kulkarni et al.]{{Girish Kulkarni$^{1}$\thanks{Email:
      kulkarni@ast.cam.ac.uk}, G\'abor Worseck$^{2}$ and Joseph
    F.~Hennawi$^{3}$} \\ $^1$Institute of Astronomy and Kavli
  Institute of Cosmology, University of Cambridge, Madingley Road,
  Cambridge CB3 0HA, UK \\ $^2$Max Planck Institute for Astronomy,
  K\"onigstuhl 17, D-69117 Heidelberg, Germany\\ $^3$Department of
  Physics, Broida Hall, UC Santa Barbara, Santa Barbara, CA 93106-9530
  USA}

\date{Accepted ---. Received ---; in original form ---}

@@

\begin{figure*}
  \begin{center}
    \includegraphics[width=\textwidth]{qsos.pdf}
    % data.py 
  \end{center}
  \caption{Quasar samples.}
  \label{fig:qsos}
\end{figure*}

\begin{table*}
  % In published version of this table, we could remove file names and add data homogenisation in Notes.
  % Add z ~ 7 qso when you add it to analysis
  % See data.tex for a version of this table that includes file names
  \caption{Quasar data sets}
  \label{tab:samples}
  \begin{tabular}{crcllrSl}
    \hline
    & ID & $z$ & Survey & Reference & Number & {Area} & Notes \\
    & & & & & of quasars & {(deg$^2$)} & \\
    \hline
    1. & 13 & 0.1--2.2 & SDSS DR7 & \citet{2006AJ....131.2766R} & 48664 & 6248.0 & \\
    2. & 15 & 0.4--2.6 & 2SLAQ SGP & \citet{2009MNRAS.392...19C} & 3663 & 64.2 & \\
    3. & 15  & 0.4--2.6 & --- NGP & \citet{2009MNRAS.392...19C} & 8153 & 127.7 & \\
    4. &  1 & 2.2--3.5 & BOSS DR9 & \citet{2013ApJ...773...14R} & 23301 & 2236.0 & \\
    5. & 13 & 3.7--4.7 & SDSS DR7 & \citet{2006AJ....131.2766R} & 1785 & 6248.0 & Restricted to $z<4.7$ \\
    6. & 17 & 4.7--5.4 & SDSS+WISE & \citet{2016ApJ...829...33Y} & 99 & 14555.0 & Overlaps with 8 for $M_{1450}<-26.73$ \\
    7. &  8 & 4.7--5.1 & SDSS DR7 & \citet{2013ApJ...768..105M} & 148 & 6248.0 & \\
    8. &  8 & 4.7--5.1 & --- Stripe 82 & \citet{2013ApJ...768..105M} & 52 & 235.0 & \\
    9. &  8 & 5.1--5.5 & --- DR7 Extended & \citet{2013ApJ...768..105M} & 28 & 6248.0 & \\
    10. & 8 & 5.1--5.4 & ---  Stripe 82 Extended & \citet{2013ApJ...768..105M} & 10 & 235.0 & \\
    11. & 6 & 3.7--4.8 & NDWFS & \citet{2011ApJ...728L..26G} & 12 & 1.71 & \\
    12. & 6 & 3.8--5.1 & DLS & \citet{2011ApJ...728L..26G} & 12 & 2.05 & \\
    13. & 7 & 4.1--6.3 & CANDELS GOODS-S & \citet{2015AA...578A..83G} & 19 & 0.047 & Rescaled selection probabilities \\
    14. & 18 & 5.8--6.4 & --- Main & \citet{2016ApJ...833..222J} & 24 & 11240.0 & \\
    15. & 18 & 5.9--6.1 & --- Overlap & \citet{2016ApJ...833..222J} & 10 & 4223.0 & \\
    16. & 18 & 5.7--6.1 & --- Stripe 82 & \citet{2016ApJ...833..222J} & 13 & 277.0 & \\
    17. & 10 & 6.0 & CFHQS Deep & \citet{2010AJ....139..906W} & 1 & 4.47 & \\
    18. & 10 & 5.9--6.4 & --- Very Wide & \citet{2010AJ....139..906W} & 16 & 494.0 & \\
    19. & 11 & 6.0--6.2 & Subaru High-$z$ Quasar & \citet{2015ApJ...798...28K} & 2 & 6.5 & \\
    20. & 19 & 6.5--7.4 & UKIDSS & \citet{2011Natur.474..616M} & 1 & 3370.0 & \\
    21. & 19 & 6.5--7.4 & UKIDSS & \citet{2007MNRAS.376L..76V} & 1 & 3370.0 & \\
    \hline
  \end{tabular}
\end{table*}

\section{Homogenised Quasar Sample}

@@

Given quasar luminosities and estimates for survey completeness, we
begin our analysis by first deriving luminosity functions binned both
in redshift and magnitude.  In this discussion all magnitudes are
absolute magnitudes at rest-frame 1450~{\AA}.

@@

  In each bin, we estimate the uncertainty in the luminosity function
  by assuming Poisson statistics \citep{1986ApJ...303..336G}.  The
  resultant binned luminosity function estimates are shown by the
  points in Figure~\ref{fig:mosaic}.  These agree quite well with the
  published luminosity functions by these surveys.

  % Do we want to give the details of how good/bad this agreement
  % actually is?

  %% Figure~\ref{fig:boss}
  %% shows the luminosity function derived in this way for the BOSS DR9
  %% colour-selected sample of 23,301 quasars, in comparison with the
  %% published luminosity function of \citet{2013ApJ...773...14R}.
  %% Figure~\ref{fig:sdss} shows the luminosity function of a sample of
  %% 15,179 quasars from the SDSS DR3 in comparison with the published
  %% luminosity function of \citet{2006AJ....131.2766R}.

@@

\begin{figure*}
  \begin{center}
    \includegraphics[width=0.7\textwidth]{evolution_individuals.pdf}
    % summary_fromFile.py 
  \end{center}
  \caption{Parameter evolution from individual fits in
    Figures~\ref{fig:mosaic}.  Parameters from BOSS quasars are seen
    to be systematically offset.}
  \label{fig:evoln}
\end{figure*}

@@

\subsection{Luminosity function estimates in redshift bins}
\label{sec:bins}

As our next step, we stop binning the quasars in luminosity, and only
bin them in redshifts.  In each redshift bin, we model the quasar
luminosity function as a double power law given by
\citep{1988MNRAS.235..935B, 1995ApJ...438..623P, 2000MNRAS.317.1014B}
\begin{equation}
  \phi(M) =
  \frac{\phi_*}{10^{0.4(\alpha+1)(M-M_*)}+10^{0.4(\beta+1)(M-M_*)}}.
  \label{eqn:dpl}
\end{equation}
Equation~(\ref{eqn:dpl}) has four free parameters: the amplitude
$\phi_*$, the ``break'' $M_*$, the faint-end slope $\beta$, and the
bright-end slope $\alpha$.  We obtain posterior probability
distributions for these parameters using MCMC \citep[e.g.,][]{jaynes}.
The joint posterior probability distribution of the model parameters
is then written as
\begin{multline}
  p(\phi_*, M_*, \alpha, \beta | \{M_i, z_i\}) \propto \\ p(\phi_*, M_*, \alpha, \beta)p(\{M_i, z_i\} | \phi_*, M_*, \alpha, \beta),
\end{multline}
where the constant of proportionality is independent of the luminosity
function parameters, and $\{M_i, z_i\}$ denotes the magnitudes and
redshifts of quasars falling in a redshift bin $[z_\mathrm{min},
  z_\mathrm{max})$.  We use a uniform prior distribution $p(\phi_*,
  M_*, \alpha, \beta)$ and assume that the likelihood
\begin{equation}
  \mathcal{L}\equiv p(\{M_i, z_i\} | \phi_*, M_*, \alpha, \beta)
\end{equation}
is given by $\phi(M)$ with suitable normalisation.  The negative
logarithm of the likelihood $S\equiv -2\ln\mathcal{L}$ can then be
written as
\begin{multline}
  S = -2\sum_{i=1}^N\ln\phi(M_i, z_i)\\+2\int_{M_\mathrm{min}}^{M_\mathrm{max}}dM\int_{z_\mathrm{min}}^{z_\mathrm{max}}dz\, \phi(M,z) f(M, z)\,\frac{dV}{dz}.
  \label{eqn:S}
\end{multline}
where $N$ is the total number of quasars in the redshift bin and the
luminosity integral in the second term on the right hand side is on
the surveyed range of $M$.

The above likelihood can also be understood as the limit of the
Poisson likelihood in luminosity and redshift bins
\citep{1983ApJ...269...35M, 2001AJ....121...54F}.  We can write the
probability of observing $n_{ij}$ quasars in the $(M_i, z_j)$ bin as
the Poisson distribution
\begin{equation}
  \mathcal{L}=\prod_{i,j}\frac{e^{-\mu_{ij}}\mu_{ij}^{n_{ij}}}{n_{ij}!},
  \label{eqn:lhood}
\end{equation}
where 
\begin{equation}
  \mu_{ij}= \int_{M_i}^{M_{i+1}}dM\int_{z_j}^{z_{j+1}}dz\, \phi(M,z) f(M, z)\,\frac{dV}{dz},
\end{equation}
is the average number of quasars expected in the $(M_i, z_j)$ bin
given the luminosity function $\phi(M,z)$.  In the limit of
infinitesimal bins, $n_{ij}=0$ or $1$, Equation~(\ref{eqn:lhood}) can
be simplified and we get Equation~(\ref{eqn:S}).

Our estimates for the luminosity function in several redshift bins are
shown in Figure~\ref{fig:mosaic}.  The median of the posterior is
shown as the best fit model.  Yellow curves show a random draw from
the posterior.  Consistent with previous studies, the double power law
model provides an excellent description of the luminosity function
model over almost the complete range of redshifts spanned by the data.
It is only in the highest redshift bin ($z=5.5$--$6.5$) that the data
seem to favour a single power law.  In this bin, we the resultant
posterior distribution of the break magnitude $M_*$ is bimodal with
favoured values at the faint ($M_*>-18$) and bright end of the data
($M_*<-30$).  We consider only the latter possibility as consistent
with the data at lower redshifts ($z<5.5$).  In light of the low
redshift data, we use restricted priors in this redshift bins: we
restrict the bright-end slope $\alpha$ to values less than $-4$.
Other paramters continue to have wide uniform priors.

As seen in Figure~\ref{fig:mosaic}, in several redshift bins the
binned quasar luminosity function seems to show a decline at the faint
end.  This is clearly seen, for example, in the 2SLAQ sample at $z<2$.
We take this to be a systematic error: the decrease in the luminosity
function is an indication of incompleteness in the quasar surveys that
has not been accounted for by the reported completeness map.  We
choose to discard such quasars from our analysis.  This behaviour is
also seen in the SDSS quasar sample at $z<1.8$ and $z\sim 4$.  The
discarded luminosity bins are shown in Figure~\ref{fig:mosaic} by open
circles.  We can see that the surviving quasar sample is in excellent
agreement with the double power law model at all redshifts.

The values of the four double power law parameters in various redshift
bins are shown in Figure~\ref{fig:evoln}.  We find interesting
evolutionary trends in each of the four parameters.  The break
luminosity $M_*$ evolves by more than eight magnitudes from redshift
$z=0$ to $7$.  The amplitude of the luminosity function $\phi_*$
evolves moderately from $z=0$ to $z\sim 3$ and then drops by six
orders of magnitude to about $10^{-12}$ at $z\sim 7$.  The bright end
slope $\alpha$ has significant scatter, but still shows a trend
towards larger negative values---that is, steeper bright-end
luminosity functions---at high redshifts.  Finally, the faint end of
the luminosity function also shows signs of increasing steepness
towards high redshifts.  Somewhat similar to the amplitude $\phi_*$,
the faint-end slope $\beta$ also shows a break at $z\sim 3$.  It stays
roughly constant up to this redshift, and the drops to increasingly
negative values at high redshifts, indicating a steeper faint end.

The behaviour of BOSS quasars in Figure~\ref{fig:evoln} is striking.
In the redshift range covered by BOSS the values of the four double
power law parameters deviate from the general trends.  The faint end
of the luminosity function of the BOSS quasars is much shallower and
the break luminosity is fainter by a magnitude.  The bright end of the
luminosity function of quasars in this sample is also much flatter
than that for other samples.  The conspicuousness of the deviation of
the BOSS sample from the general trends makes it unlikely that the
luminosity function evolution indicated by BOSS in the redshift range
$z=2$--$3$ is physical.  Indeed, in this redshift range BOSS quasar
selection is known to be affected by a colour bias
\citep{2011ApJ...728...23W}, as the quasar colour selection is
contaminated by a population of blue stars.  We therefore deem the
BOSS sample as affected by this systematic error.

The open circles in Figure~\ref{fig:evoln} show the results of a
similar analysis of the quasar UV luminosity function by
\citep{2017MNRAS.466.1160M}.  Evidently, our results are quite
different from the results of these authors.  While their analysis
generally also reveals a steep faint end slop, the break luminosity
evolution is dramatically different.  In their analysis the break
luminosity tends to remain roughly constant $M_*\sim -24$ from $z=0$
to $7$, but this does not seem to be indicated by the data shown in
Figure~\ref{fig:mosaic}.  A possible reason behind their significanly
different results could be a bias introduced by binning in
luminosities and redshifts while fitting the luminosity function model.

@@

\subsection{Luminosity function estimate without binning}

@@

In the local source approximation, Equation~(\ref{eqn:flux}) becomes
\begin{equation}
  j(\nu_0, z_0) = \frac{1}{4\pi}\lambda_\mathrm{mfp}(\nu_0, z_0)\epsilon(\nu_0, z_0),
\end{equation}
where $\lambda_\mathrm{mfp}$ is the mean free path.  The hydrogen
photoionization rate is given by
\begin{equation}
  \Gamma_\mathrm{HI}=4\pi\int_{\nu_{912}}^\infty d\nu \frac{j_\nu}{h\nu} \sigma(\nu),
  \label{eqn:gammapi}
\end{equation}
where $\sigma$ is the ionization cross-section
\begin{equation}
  \sigma(\nu) = \sigma_0\left(\frac{\nu}{\nu_{912}}\right)^{-3},
  \label{eqn:sigma}
\end{equation}
with $\sigma_0=6.3\times 10^{-18}$~cm$^2$ \citep{2006agna.book.....O}.
In the redshift range $z=2.3$--$5.5$, the mean free path is measured
to be \citep{2014MNRAS.445.1745W}
\begin{equation}
  \lambda_\mathrm{mfp}(\nu, z)= \lambda_0\left(\frac{1+z}{5}\right)^{-5.4}\left(\frac{\nu}{\nu_{912}}\right)^{1.5},
  \label{eqn:mfp}
\end{equation}
where the frequency dependence comes from the assumed column density
distribution $f(N_\mathrm{HI})\propto N_\mathrm{HI}^{-1.5}$.  We
extrapolate the redshift dependence in Equation~(\ref{eqn:mfp}) at
$z<2.3$ and $z>5.5$.  Combining Equations~(\ref{eqn:epsilon_freq}),
(\ref{eqn:sigma}), and (\ref{eqn:mfp}), Equation~(\ref{eqn:gammapi})
gives
\begin{multline}
  \Gamma_\mathrm{HI}=4.6\times 10^{-13} \mathrm{s}^{-1} \left(\frac{\epsilon_{912}}{10^{24}\mathrm{erg s^{-1} Hz^{-1} cMpc^{-3}}}\right)\\
  \times\left(\frac{1}{1.5-\alpha}\right)\left(\frac{1+z}{5}\right)^{-2.4}.
\end{multline}
Note that this equation refers to the \emph{comoving} emissivity.  The
local source approximation works well for $z>3$, as can be seen in
Figure~\ref{fig:gammapi}, where the points from the individual fits
are in much better agreement with the curves from the global model.
At lower redshift, the photoionisation rate in the individual models
is much higher than that in the global model, because the individual
models overestimate the flux at these redshifts.

@@

\section{Hydrogen and helium reionization}

@@

\section{Photon underproduction}

@@

It is a pleasure to acknowledge useful discussions with James Aird,
Manda Banerji, Tirthankar Roy Choudhury, George Efstathiou, Xiaohui
Fan, Andrea Ferrara, Prakash Gaikwad, Martin Haehnelt, Paul Hewett,
Sergey Koposov, Donald Lynden-Bell, Roberto Maiolino, Richard McMahon,
Ewald Puchwein, Gordon Richards, Alberto Rorai, and Stephen Warren.
GK acknowledges support from ERC Advanced Grant 320596 `The Emergence
of Structure During the Epoch of Reionization'. \gk{Perhaps we need to
  specially thank people who shared data with us.  Gabor, Joe: do you
  need to add anything to acknowledgements?}

>>>>

\author[Kulkarni et al.]{{Girish Kulkarni$^{1}$\thanks{Email:
      kulkarni@ast.cam.ac.uk}, G\'abor Worseck$^{2}$ and Joseph
    F.~Hennawi$^{3}$} \\ $^1$Institute of Astronomy and Kavli
  Institute of Cosmology, University of Cambridge, Madingley Road,
  Cambridge CB3 0HA, UK \\ $^2$Max Planck Institute for Astronomy,
  K\"onigstuhl 17, D-69117 Heidelberg, Germany\\ $^3$Department of
  Physics, Broida Hall, UC Santa Barbara, Santa Barbara, CA 93106-9530
  USA}
% Update Gabor's address to Potsdam
\date{Accepted ---. Received ---; in original form ---}

@@

\begin{figure*}
  \begin{center}
    \includegraphics[width=\textwidth]{qsos.pdf}
    % data.py 
  \end{center}
  \caption{Redshift distribution of the XX quasars considered in this
    analysis.  Shown here are the observed quasar numbers, without
    correcting for incompleteness.  Further details on each of these
    data sets are in Table~\ref{tab:samples} and
    Section~\ref{sec:sample}.}
  \label{fig:qsos}
\end{figure*}

\begin{table*}
  % In published version of this table, we could remove file names and add data homogenisation in Notes.
  % Add z ~ 7 qso when you add it to analysis
  % See data.tex for a version of this table that includes file names
  % Should we mark Giallongo, Mortlock, and Banados entries to say that these are sometimes excluded from analysis?
  % Remove notes?
  \caption{Quasar data sets}
  \label{tab:samples}
  \begin{tabular}{crcllrSl}
    \hline
    & ID & $z$ & Survey & Reference & Number & {Area} & Notes \\
    & & & & & of quasars & {(deg$^2$)} & \\
    \hline
    1. & 13 & 0.1--2.2 & SDSS DR7 & \citet{2006AJ....131.2766R} & 48664 & 6248.0 & \\
    2. & 15 & 0.4--2.6 & 2SLAQ SGP & \citet{2009MNRAS.392...19C} & 3663 & 64.2 & \\
    3. & 15  & 0.4--2.6 & --- NGP & \citet{2009MNRAS.392...19C} & 8153 & 127.7 & \\
    4. &  1 & 2.2--3.5 & BOSS DR9 & \citet{2013ApJ...773...14R} & 23301 & 2236.0 & \\
    5. & 13 & 3.7--4.7 & SDSS DR7 & \citet{2006AJ....131.2766R} & 1785 & 6248.0 & Restricted to $z<4.7$ \\
    6. & 17 & 4.7--5.4 & SDSS+WISE & \citet{2016ApJ...829...33Y} & 99 & 14555.0 & Overlaps with 8 for $M_{1450}<-26.73$ \\
    7. &  8 & 4.7--5.1 & SDSS DR7 & \citet{2013ApJ...768..105M} & 148 & 6248.0 & \\
    8. &  8 & 4.7--5.1 & --- Stripe 82 & \citet{2013ApJ...768..105M} & 52 & 235.0 & \\
    9. &  8 & 5.1--5.5 & --- DR7 Extended & \citet{2013ApJ...768..105M} & 28 & 6248.0 & \\
    10. & 8 & 5.1--5.4 & ---  Stripe 82 Extended & \citet{2013ApJ...768..105M} & 10 & 235.0 & \\
    11. & 6 & 3.7--4.8 & NDWFS & \citet{2011ApJ...728L..26G} & 12 & 1.71 & \\
    12. & 6 & 3.8--5.1 & DLS & \citet{2011ApJ...728L..26G} & 12 & 2.05 & \\
    13. & 7 & 4.1--6.3 & CANDELS GOODS-S & \citet{2015AA...578A..83G} & 19 & 0.047 & Rescaled selection probabilities \\
    14. & 18 & 5.8--6.4 & --- Main & \citet{2016ApJ...833..222J} & 24 & 11240.0 & \\
    15. & 18 & 5.9--6.1 & --- Overlap & \citet{2016ApJ...833..222J} & 10 & 4223.0 & \\
    16. & 18 & 5.7--6.1 & --- Stripe 82 & \citet{2016ApJ...833..222J} & 13 & 277.0 & \\
    17. & 10 & 6.0 & CFHQS Deep & \citet{2010AJ....139..906W} & 1 & 4.47 & \\
    18. & 10 & 5.9--6.4 & --- Very Wide & \citet{2010AJ....139..906W} & 16 & 494.0 & \\
    19. & 11 & 6.0--6.2 & Subaru High-$z$ Quasar & \citet{2015ApJ...798...28K} & 2 & 6.5 & \\
    20. & 19 & 6.5--7.4 & UKIDSS & \citet{2011Natur.474..616M} & 1 & 3370.0 & \\
    21. & 19 & 6.5--7.4 & UKIDSS & \citet{2007MNRAS.376L..76V} & 1 & 3370.0 & \\
    22. & 20 & 6.5--7.4 & ALLWISE+UKIDSS+DECaLS & \citet{2018Natur.553..473B} & 1 & 2500.0 & \\
    \hline
  \end{tabular}
\end{table*}

\section{Homogenised Quasar Sample}
\label{sec:sample}

@@

Given quasar luminosities and estimates for survey completeness, we
are in a position to estimate the UV luminosity function of AGN.  We
begin our analysis by first deriving luminosity functions in bins of
redshift and magnitude.  In this discussion all magnitudes are
absolute magnitudes at rest-frame 1450~{\AA}.

@@

  In each bin, we estimate the uncertainty in the luminosity function
  by assuming Poisson statistics \citep{1986ApJ...303..336G} for the
  number of quasars in each bin.  The resultant binned luminosity
  function estimates are shown by the points in
  Figure~\ref{fig:mosaic}.

  As seen in Figure~\ref{fig:mosaic}, the distribution of luminosity
  function values in each redshift bin are suggestive of a double
  power law form.  We will fit such a form below.  However, we also
  see in several redshift bins of Figure~\ref{fig:mosaic} that the
  luminosity function seems to show a decline at the faint end.  This
  is clearly seen, for example, in the 2SLAQ sample at $z<2$ and in
  the SDSS quasar sample at $z<1.8$ and $z\sim 4$.  We take this to be
  a systematic error.  The decrease in the luminosity function is an
  indication of further low-level incompleteness in the quasar surveys
  that has not been captured by the reported completeness values.  We
  choose to discard such bins from our analysis.  The discarded
  luminosity bins are shown in Figure~\ref{fig:mosaic} by open
  circles.

@@

\begin{figure*}
  \begin{center}
    \includegraphics[width=0.7\textwidth]{evolution_individuals.pdf}
    % summary_fromFile.py 
  \end{center}
  \caption{Evolution of the four double power law parameters from the
    redshift bins shown in Figure~\ref{fig:mosaic}.  Vertical error
    bars show one-sigma (68.26\%) uncertainties, whereas horizontal
    error bars show widths of the redshift bins.  We identify the
    general evolutionary trends of each of these parameters from the
    bins shown by the filled symbols.  The open symbols show bins that
    appear to be offset from these trends, likely due to unknown
    systematic errors.  The open circles at $z\sim 2$--$4$ show the
    BOSS sample, while the bins at $z < 0.6$ contain AGN from the SDSS
    and 2SLAQ data sets.  See Section~\ref{sec:bins} for further
    details.}
  \label{fig:evoln}
\end{figure*}

@@

\subsection{Double power law fits}
\label{sec:bins}

As our next step, we combine magnitude bins by binning the quasars
only in redshifts.  This allows us to estimate the luminosity function
at each redshift.  In each redshift bin, we model the quasar
luminosity function as a double power law given by
\citep{1988MNRAS.235..935B, 1995ApJ...438..623P, 2000MNRAS.317.1014B}
\begin{equation}
  \phi(M) =
  \frac{\phi_*}{10^{0.4(\alpha+1)(M-M_*)}+10^{0.4(\beta+1)(M-M_*)}}.
  \label{eqn:dpl}
\end{equation}
Equation~(\ref{eqn:dpl}) has four free parameters: the amplitude
$\phi_*$, the break magnitude $M_*$, the faint-end slope $\beta$, and
the bright-end slope $\alpha$.  By assuming broad, uniform priors, we
obtain posterior probability distributions for these parameters using
MCMC \citep[e.g.,][]{jaynes}.  The joint posterior probability
distribution of the model parameters is then written as
\begin{multline}
  p(\phi_*, M_*, \alpha, \beta | \{M_i, z_i\}) \propto \\ p(\phi_*, M_*,
  \alpha, \beta)p(\{M_i, z_i\} | \phi_*, M_*, \alpha, \beta),
\end{multline}
where the constant of proportionality is independent of the luminosity
function parameters, and $\{M_i, z_i\}$ denotes the magnitudes and
redshifts of quasars falling in a redshift bin $[z_\mathrm{min},
  z_\mathrm{max})$.  We use a uniform prior distribution $p(\phi_*,
  M_*, \alpha, \beta)$ and assume that the likelihood
\begin{equation}
  \mathcal{L}\equiv p(\{M_i, z_i\} | \phi_*, M_*, \alpha, \beta)
\end{equation}
is given by $\phi(M)$ with suitable normalisation.  The negative
logarithm of the likelihood $S\equiv -2\ln\mathcal{L}$ can then be
written as
\begin{multline}
  S = -2\sum_{i=1}^N\ln\phi(M_i, z_i)\\+2\int_{M_\mathrm{min}}^{M_\mathrm{max}}dM
  \int_{z_\mathrm{min}}^{z_\mathrm{max}}dz\, \phi(M,z) f(M, z)\,\frac{dV}{dz}.
  \label{eqn:S}
\end{multline}
where $N$ is the total number of quasars in the redshift bin and the
luminosity integral in the second term on the right hand side is on
the surveyed range of $M$.  We use the \texttt{emcee} code
\citep{2013PASP..125..306F} for MCMC.

The above likelihood can also be understood as the limit of the
Poisson likelihood in luminosity and redshift bins
\citep{1983ApJ...269...35M, 2001AJ....121...54F}.  We can write the
probability of observing $n_{ij}$ quasars in the $(M_i, z_j)$ bin as
the Poisson distribution
\begin{equation}
  \mathcal{L}=\prod_{i,j}\frac{e^{-\mu_{ij}}\mu_{ij}^{n_{ij}}}{n_{ij}!},
  \label{eqn:lhood}
\end{equation}
where 
\begin{equation}
  \mu_{ij}= \int_{M_i}^{M_{i+1}}dM\int_{z_j}^{z_{j+1}}dz\, \phi(M,z) f(M, z)
  \,\frac{dV}{dz},
\end{equation}
is the average number of quasars expected in the $(M_i, z_j)$ bin
given the luminosity function $\phi(M,z)$.  In the limit of
infinitesimal bins, $n_{ij}=0$ or $1$, and Equation~(\ref{eqn:lhood})
can be simplified to obtain Equation~(\ref{eqn:S}).

Our estimates for the double power law luminosity function are shown
in Figure~\ref{fig:mosaic} for 25 redshift bins.  The posterior median
is shown as the best-fit model at each redshift, while the grey shaded
region shows the one-sigma (68.26\%) uncertainty.  Consistent with
previous studies, the double power law model provides an excellent
description of the luminosity function model over almost the complete
range of redshifts spanned by the data.  It is only in the highest
redshift bin ($z=5.5$--$6.5$) that the data seem to favour a single
power law.  In this bin, the resultant posterior distribution of the
break magnitude $M_*$ is bimodal with favoured values at the faint
($M_*>-18$) and bright end of the data ($M_*<-30$).  While in the
literature the $z\sim 6$ AGN have been assumed to lie on the bright
end of the luminosity function \citep[e.g.,][]{2016ApJ...833..222J}, a
comparison with the luminosity function at lower redshifts ($z<5.5$)
suggests that these AGN should instead be understood to describe the
faint-end of a double power law.  Therefore, in light of the
low-redshift data, we use restricted priors in this redshift bin: we
restrict the bright-end slope $\alpha$ to values less than $-4$, which
is equivalent to forcing $M_*$ to be at the bright end of the data.
Other parameters continue to have wide uniform priors.  This also
illustrates the importance of analysing the high-redshift AGN
population in a wider context of AGN data.

The redshift evolution of the four double power law parameters is
shown in Figure~\ref{fig:evoln}.  These values have been tabulated in
Table~\ref{tab:bins}.  We find interesting evolutionary trends in each
of the four parameters.  The break luminosity $M_*$ evolves by more
than eight magnitudes from redshift $z=0$ to $7$.  The amplitude of
the luminosity function $\phi_*$ evolves moderately from $z=0$ to
$z\sim 3$ and then drops by six orders of magnitude to about
$10^{-12}$ at $z\sim 7$.  The bright end slope $\alpha$ has
significant scatter, but still shows a trend towards larger negative
values, i.e., steeper luminosity function bright ends, at high
redshifts.  Finally, the faint end of the luminosity function also
shows signs of increasing steepness towards high redshifts.  Somewhat
similar to the amplitude $\phi_*$, the faint-end slope $\beta$ also
shows a break at $z\sim 3$.  It stays roughly constant up to this
redshift, and the drops to increasingly negative values of close to
$-2.5$ at high redshifts, indicating a steeper faint end.

Figure~\ref{fig:evoln} reveals further potentially systematic errors.
The behaviour of BOSS quasars, shown by open circles in
Figure~\ref{fig:evoln} at redshifts $z=2$--$4$ is striking.  In this
redshift range covered by BOSS the values of the four double power law
parameters deviate from the general trends.  The faint end of the
luminosity function of the BOSS quasars is much shallower and the
break luminosity is fainter by a magnitude compared to the general
trend expected from a smooth evolution.  The bright end of the
luminosity function of quasars in this sample is also much flatter
than that for other samples.  The conspicuousness of the deviation of
the BOSS sample from the general trends makes it unlikely that the
luminosity function evolution indicated by BOSS in the redshift range
$z=2$--$3$ is physical.  Indeed, in this redshift range BOSS quasar
selection is known to be affected by a colour bias
\citep{2011ApJ...728...23W}, which can make estimating completeness
difficult \citep{2006AJ....131.2766R}.  \gk{Gabor, could you please
  check if I have written this paragraph OK?}

@@

%--------------------

\subsection{Evolution of the luminosity function}

@@

In the local source approximation, Equation~(\ref{eqn:flux}) becomes
\begin{equation}
  j(\nu_0, z_0) = \frac{1}{4\pi}\lambda_\mathrm{mfp}(\nu_0, z_0)\epsilon(\nu_0, z_0),
\end{equation}
where $\lambda_\mathrm{mfp}$ is the mean free path.  The hydrogen
photoionization rate is given by
\begin{equation}
  \Gamma_\mathrm{HI}=4\pi\int_{\nu_{912}}^\infty d\nu \frac{j_\nu}{h\nu} \sigma(\nu),
  \label{eqn:gammapi}
\end{equation}
where $\sigma$ is the ionization cross-section
\begin{equation}
  \sigma(\nu) = \sigma_0\left(\frac{\nu}{\nu_{912}}\right)^{-3},
  \label{eqn:sigma}
\end{equation}
with $\sigma_0=6.3\times 10^{-18}$~cm$^2$ \citep{2006agna.book.....O}.
In the redshift range $z=2.3$--$5.5$, the mean free path is measured
to be \citep{2014MNRAS.445.1745W}
\begin{equation}
  \lambda_\mathrm{mfp}(\nu, z)= \lambda_0\left(\frac{1+z}{5}\right)^{-5.4}\left(\frac{\nu}{\nu_{912}}\right)^{1.5},
  \label{eqn:mfp}
\end{equation}
where the frequency dependence comes from the assumed column density
distribution $f(N_\mathrm{HI})\propto N_\mathrm{HI}^{-1.5}$.  We
extrapolate the redshift dependence in Equation~(\ref{eqn:mfp}) at
$z<2.3$ and $z>5.5$.  Combining Equations~(\ref{eqn:epsilon_freq}),
(\ref{eqn:sigma}), and (\ref{eqn:mfp}), Equation~(\ref{eqn:gammapi})
gives
\begin{multline}
  \Gamma_\mathrm{HI}=4.6\times 10^{-13} \mathrm{s}^{-1} \left(\frac{\epsilon_{912}}{10^{24}\mathrm{erg\, s^{-1}\, Hz^{-1}\, cMpc^{-3}}}\right)\\
  \times\left(\frac{1}{1.5-\alpha}\right)\left(\frac{1+z}{5}\right)^{-2.4}.
\end{multline}
Note that this equation refers to the \emph{comoving} emissivity.  The
local source approximation works well for $z>3$, as can be seen in
Figure~\ref{fig:gammapi}, where the points from the individual fits
are in much better agreement with the curves from the global model.
At lower redshift, the photoionisation rate in the individual models
is much higher than that in the global model, because the individual
models overestimate the flux at these redshifts.

@@

\subsection{Photon underproduction at $z=0$?}

\section{Helium reionization}

@@

@@

We thank Eilat Glikman, Linhua Jiang, Nobunari Kashi\-kawa, Ian
McGreer, Nick Ross, Chris Willott, and Jinyi Yang for sharing data and
especially for sharing their quasar selection functions.  It is a
pleasure to acknowledge useful discussions with James Aird, Eduardo
Ba\~nados, Manda Banerji, Tirthankar Roy Choudhury, George Efstathiou,
Xiaohui Fan, Andrea Ferrara, Prakash Gaikwad, Martin Haehnelt, Paul
Hewett, David Hogg, Vikram Khaire, Sergey Koposov, Donald Lynden-Bell,
Roberto Maiolino, Richard McMahon, Daniel Mortlock, Ewald Puchwein,
Gordon Richards, Alberto Rorai, Bram Venemans and Stephen Warren.  GK
acknowledges support from ERC Advanced Grant 320596 `The Emergence of
Structure During the Epoch of Reionization'. \gk{Joe, do you need to
  add anything to acknowledgements?}
